Lyman-$\alpha$ photons through rotating outflows

### 2018-09-15 paper/paper.tex
update

As input parameters we use $\tauh=\{10^5, 10^6, 10^7\}$,
$\vout=\{5,25,50\}\;\kms$ and $\vrot=\{0,50,100\}\;\kms$, for a total of
$27$ models with all the possible parameter combinations.
The values for the outflow velocity are lower than values commonly used in the
literature to allow for an interplay between the two kinematic features.
Nevertheless, this range of parameters reproduces the main standard
deviation and skewness  of high redshift LAES in recent observations
(\label{subsec:kinematic}).

>>>>

As input parameters we use $\tauh=\{10^5, 10^6, 10^7\}$,
$\vout=\{5,25,50\}\;\kms$ and $\vrot=\{0,50,100\}\;\kms$, for a total of
$27$ models with all the possible parameter combinations.
The values for the outflow velocity are lower than values commonly used in the
literature to allow for an interplay between the two kinematic features.
This range of parameters also produce emission lines with standard deviation and
skewness in the same range to those of high redshift LAES in recent
observations \citep{2017A&A...606A..12H}.

### 2018-09-15 paper/paper_old.tex
update

%=========================================================================
%		FRONT MATTER
%=========================================================================
\title[Lyman-$\alpha$ photons through rotating
  outflows]{Lyman-$\alpha$ photons through rotating outflows}  
\author[M.C. Remolina-Gutierrez \& J.E. Forero-Romero]{
  Maria Camila Remolina-Guti\'errez$^{1}$
  \thanks{mc.remolina197@uniandes.edu.co} \&
  Jaime E. Forero-Romero $^{1}$
  \thanks{je.forero@uniandes.edu.co}\\
  %%
  $^{1}$ Departamento de F\'isica, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia \\
}

@@

\begin{abstract}
Outflows and rotation are two ubiquitous kinematic features in the gas
kinematics of galaxies. 
Here we introduce a semi-analytic model to quantify how rotating
outflows impact the morphology of the Lyman-$\alpha$ emission line.   
The model is contrasted against Monte Carlo radiative transfer
simulations of outflowing gas with additional solid body rotation.
We explore a range of neutral Hydrogen optical depth of
$10^5\leq\tauh\leq10^7$, rotational velocity $0\leq \vrot/\kms \leq
100$ and outflow velocity $0\leq \vout/\kms\leq 50$.  
We find three consequences of rotation.
First, it introduces a dependency with viewing angle; second it
broadens the line and third it increases the flux at the
line's center.
For all viewing angles, the semi-analytic model reproduces the
radiative transfer results for the line width and flux change at the
line's center within a $7\%$ and $50\%$ precision for an optical depth of
$\tauh=10^5$, respectively, and within $2\%$ and $1\%$ for an optical
depth of $\tauh=10^7$.
Using this model we also show that the peaks of integrated spectra
taken from opposite sides of an edge-on rotating gas distribution
should have a separation of $\frac{1}{2}\vrot$. 
The semi-analytic model presented here is a convenient tool to
introduce rotational kinematics as a post-processing step of idealized
Monte Carlo simulations; it provides a framework to interpret \lya
spectra in systems where rotation is expected or directly measured
through kinematic maps.  
\end{abstract}

@@

Recent advances in instrumentation have revealed the presence of gas
rotation on vastly different physical scales.
For instance, spatially resolved spectra on compact dwarf galaxies
have measured clear signs of gas showing pure rotation kinematics
\citep{2015A&A...577A..21C,2017A&A...600A.125C} and the recent mapping of
high redshift circumgalactic regions have also revealed kinematic
evidence for large scale rotation \citep{2018MNRAS.473.3907A}.
Systems with star formation, neutral gas and low dust
contents can produce a \lya emission line \citep{PartridgePeebles}
motivating the observational work to phenomenologically link tracers
of galaxy rotation such as H$\alpha$ to \lya spectra
\citep[e.g.][]{Herenz2016}.

@@

However, the gas dynamics in Lyman Alpha Emitter (LAE) galaxies are
more complex than pure rotation.
In many observations the \lya line profile has a single peak
redwards from the line's center, in other cases there is a double peak
but the peak on the red side is stronger
\citep[e.g.][]{2010ApJ...717..289S,Erb14,Trainor16}.   
These features have been explained as the consequence of multiple
\lya photon scatterings through a homogeneous outflowing shell of
neutral Hydrogen
\citep{Verhamme06,Orsi12,2015ApJ...812..123G}.

Nevertheless, a study of the combined effects of outflows and
rotation has not been presented in the literature.
Here we report on such a study with the main aim of quantifying the
validity of the semi-analytic approximation presented by
\cite{Garavito14} in the case where outflows are also present. 
We investigate a simplified geometrical configuration corresponding to a
spherical gas cloud with symmetrical radial outflows and solid body
rotation and contrast the semi-analytic model against the results of a
Monte-Carlo radiative transfer code.

@@

Throughout the paper we use a thermal velocity for a neutral Hydrogen
gas of \vth $= 12.86$ \kms, which corresponds to a temperature of
$T=10^4$ K.

@@

@@

We use CLARA \citep{CLARA}, a Monte Carlo code that follows the
propagation of individual photons through a neutral Hydrogen medium
characterized by its temperature, velocity field and global optical depth.
The code assumes an homogeneous density throughout the simulated volume.
In the current implementation we neglect the influence of dust.
Our basic model is an spherical distribution of neutral hydrogen,
an approximation commonly used in the literature that explains a
wide variety of observational features
\citep{Ahn03,Verhamme06,Dijkstra06}.

We use a velocity field that captures both outflows and rotation.
Outflows are described by a Hubble-like radial velocity profile with
the speed increasing linearly with the radial coordinate; this model
is fully characterized by $v_{\rm out}$, the radial velocity at the
sphere's surface. 
Rotation follows a solid body rotation profile fully characterized by
$v_{\rm rot}$ the linear velocity at the sphere's surface.

The total velocity field is the superposition of rotation and
outflows.
The cartesian components take the following form:

@@

\begin{equation}
	v_{z}=\frac{z}{R}\vout,
	\label{eq:vz}
\end{equation}
%
where $x$, $y$ and $z$ are the cartesian position coordinates with the
origin at the sphere's center, $R$ is the radius of the sphere and the
direction of the angular velocity vector is the $z$ axis.

For each run we follow $10^5$ individual photons generated from the
origin at the \lya line's center as they propagate
through the volume and finally escape.
We store the frequency and propagation direction for each photon
at its last scattering.

As input parameters we use $\tauh=\{10^5, 10^6, 10^7\}$,
$\vout=\{5,25,50\}\;\kms$ and $\vrot=\{0,50,100\}\;\kms$, for a total of
$27$ models with all the possible parameter combinations.
The values for the outflow velocity are lower than values commonly used in the
literature to allow for an interplay between the two kinematic features.
This range of parameters also produce emission lines with standard deviation and
skewness in the same range to those of high redshift LAES in recent
observations \citep{2017A&A...606A..12H}.

We also define the viewing angle, $\theta$, as the angle
between the rotation axis and the line of sight of a potential
observer.

\cite{Garavito14} presented an analytical model that
accounts for the effects of pure rotation on the
\lya line morphology. 
The basic assumption of the model is that each differential surface
element on the sphere Doppler shifts the photons that it emits.
In this paper we introduce this ansatz by post-processing the results
of the outflows simulations without rotation.
The frequency of each photon is Doppler shifted as follows

\begin{equation}
x' = x + \frac{\vec{v}_{\rm rot}\cdot \hatk}{\vth}
\label{eq:shift_x}
\end{equation}
%
where $x'$ is the photon's new adimensional frequency, $x$ is the photon's
frequency after being processed only by the outflow, \vrot is the 
rotational velocity at the point of escape of the photon, \hatk is
the photon's direction of propagation and \vth is the thermal
velocity of the sphere. 
We fix \vth throughout the paper. This factor is a multiplying
constant that could be changed without running new Monte-Carlo
simulations, as the code works with the adimensional frequency $x$.

This semi-analytic model allows us to produce new \lya spectra from
the outflow-only results and compare them with the full radiative
transfer solution including both outflows and rotation.

@@

Figure \ref{fig:doppler_shift} summarizes the most important trends
from rhw RT simulations.
In the left side, the six panels correspond to $\tau=10^6$ and a viewing angle of
$\theta =90^{\circ}$, that is, perpendicular to the rotation axis of the
galaxy. 
In every panel the thin black line corresponds to the pure outflow
solution, i.e. without rotation. 
From top to bottom we see the effect of increasing the outflow
velocity, which is the expected increasing asymmetry towards the red
peak.

@@

The thick gray line corresponds to the pure outflow solution
with the Doppler shift added to model rotation's influence.
At $\tauh=10^6$ the Doppler shift does a good job at capturing the broad
morphological features introduced by rotation: the angle dependence,
the broadening and the intensity increase at the line's center.

In the right side of Figure \ref{fig:doppler_shift} we show the same results as in the left one, but for a viewing angle of $\theta =
0^{\circ}$, that is parallel to the rotation axis. 
In this case we confirm the result presented by \cite{Garavito14},
namely that pure rotation introduces a strong dependence with 
viewing angle, a trend that we find also holds for rotation mixed with
outflows.

The quality of the results from the Doppler shift improves for higher
\tauh values. 
In the Appendix we show the same plots as Figure
\ref{fig:doppler_shift}, there it is evident that for $\tauh=10^5$ the
results are not as good as they are for $\tauh=10^6$, and that for
$\tauh=10^7$ the Doppler shift provides a remarkable good
approximation.

\subsection{Quantitative Trends}
\label{sec:quantitative}

After finding the qualitative influence of the different parameters we
move onto a quantitative study.
To do this we summarize the line morphology by four different
scalars: standard deviation (STD), skewness (SKW), bimodality
(BI) and valley/peak ratio.
These quantities are defined by the following equations \citep{kokoska1999}:

@@

\subsection{The semi-analytic model as gaussian smoothing}

The effects of the semi-analytic model on the pure outflowing spectra
are similar to the expected results from a gaussian smoothing, 
such smoothing can also be a natural consequence of the intruments
used to measure the LAE spectra.
As a consequence, natural experimental artifacts could be mistaken as
an indication for rotation.

To understand to what extent the effects of rotation can be seen as a
simple smoothing, we model the effects of gaussian smoothing with a
similar ansatz as  the semi-analytic model by changing the frequency
of each photon by a Gaussian random variable centered on zero and
standard deviation $\sigma_x$: \begin{equation}
x^{\prime} = x + \mathcal{N}(0,\sigma_x^2).
\end{equation}

We find that the results of the semi-analytic model on the total
spectra at a given angle $\theta$ can be approximated using
$\sigma_x = (1/2)\times(\vrot/\vth)\times \sin\theta$. 
This approximation reproduces within a few porcent all quantities
measured in Section \ref{sec:quantitative} for the semi-analytic
model, it only works well for $\tau=10^6$ and $\tau=10^7$.

This result can be used as an order of magnitude estimate of the mimum
espectral resolution required to observe the effects of a rotational
velocity of $\vrot$. 
The Space Telescope Imaging Spectrograph (STIS) on the Hubble Space
Telescope (HST) can provide resolutions around 20 \kms
\citep{2015PASA...32...27H}, which could allow for the detection of
rotation features in nearby galaxies with at least $40$ \kms and a
rotation axis perpendicular to the line of sight.  
Other instruments such as the Cosmic Origins Spectrograph (COS) have
varying resolutions between $20$\kms and $200$\kms according to the
angular size of the source \citep{2018A&A...616A..60O}.

\subsection{\lya Kinematic Maps}
\label{subsec:kinematic}

Current observational facilities have the capability of spatially
resolving the extent of a LAE.
For instance \cite{Prescott14} presented observational results of a
Doppler shift when taking spectra at two opposite sides of a large
($\approx 80$kpc) LAE.
In more recent work \cite{2018MNRAS.473.3907A} mapped \lya emission
around a quasar.
In this case one could measure the displacement between the
peaks in the two different spectra. 
The interest of this test is that the location of the peaks only
depends on the kinematic properties of the sources and does not
change by convolution with a gaussian window function due to limited
instrumental resolution, as discussed in the previous section.

@@

In this toy model the distance between the peaks of the
receding/approaching spectra is close to $\sim 25\ \kms$, which is
a fourth of the naively expected value of $2\vrot=100\ \kms$,
due to the fact that only a small fraction of the photons are emitted
at the extreme of the galaxy having the maximum rotational velocity of
$50\ \kms$. 
Although it is a good approximation to think the rotating spectra by a
sum of Doppler shifts, the peak of the spectra is also weighted by the
amount of mass with a given line-of-sight velocity.

Spectrographs like the Multi Unit Spectroscopic Explorer (MUSE) could
obtain kinematic information from large samples of LAEs to build
velocity maps in Ly$\alpha$.  
This could be a natural extension of the work reported by
\cite{Herenz2016} on the velocity maps of several LARS (Lyman Alpha
Reference Sample) galaxies.  
The interpretation of such data could take into account the insights
and trends we have presented in this paper.

@@

In this paper we explore, for the first time in the literature,
the results of a model for the emergent \lya line from rotating
outflows. 
We the a semi-analytic model first presented by \cite{Garavito14}
to capture the main effects of rotation and confront it against
results from Monte-Carlo radiative transfer simulations. 
The semi-analytic model only takes into account the Doppler shift computed
as product of quantities at the surface of last scattering, namely
$\vec{v}_{\rm rot}\cdot \hatk$, where $\vec{v}_{\rm rot}$ is the
velocity due to rotation and \hatk is the direction of the photon's
propagation.

To address the first question we posed in the introduction (\emph{What
  is the expected imprint of rotation on a resonant emission line such
  as the \lya line?}) we find from the full RT simulations that the
effects of rotation on the \lya line morphology are:

\begin{itemize}
  \item Inducing a dependency on the viewing angle.
  \item Broadening the line.
  \item Increasing the intensity at the line's center.
\end{itemize}
%
All these effects can be qualitatively explained by the proposed
semi-analytic model. 
Quantitatively speaking the semi-analytic model provides a
satisfactory answer for a neutral Hydrogen optical depth equal or
larger than $10^6$.

Addressing the second question we posed in the introduction (\emph{To what
extent is it possible to constrain rotational kinematics from the \lya
emission line?})  we show that the most straightforward approach
is measuring \lya spectra from two different regions of a
galaxy to detect approaching/receeding gas motions
\citep{Prescott14,2018MNRAS.473.3907A}. 
In that case, we also show that the distances between these two peaks 
is four times shorter than the naively expected value of
$2\times\vrot$.  
This difference is produced by the different geometric weights at the
emitting surface.    
Here the semi-analytic model also provides the correct quantitative
insight.

To summarize, our works shows that the Doppler Shift offers an
easy-to-implement approximation to explore such influence into already
existing radiative transfer simulation simulations that include
outflows, providing a versatile tool to interpret current and future \lya
kinematics maps  \citep[e.g][]{2018MNRAS.473.3907A}.

@@

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.49\textwidth]{doppler_shift_logtau5_theta90}
    \includegraphics[width=0.49\textwidth]{doppler_shift_logtau7_theta90}
  \end{center}
  \caption{\textbf{Qualitative trends of changing outflow and
      rotational velocity.}
    Same layout as Figure \ref{fig:doppler_shift}. On the left:
    $\tauh=10^5$ and $\theta=90^\circ$; on the right:     this time  $\tauh=10^7$ and $\theta=90^\circ$.}
\end{figure*}

>>>>

%=========================================================================
%		FRONT MATTER
%=========================================================================
\title[Lyman-α photons through rotating outflows]{Lyman-$\alpha$ photons through rotating outflows}
\author[M.C. Remolina-Gutierrez \& J.E. Forero-Romero]{
  Maria Camila Remolina-Guti\'errez$^{1}$
  \thanks{mc.remolina197@uniandes.edu.co} \&
  Jaime E. Forero-Romero $^{1}$
  \thanks{je.forero@uniandes.edu.co}\\
  %%
  $^{1}$ Departamento de F\'isica, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia \\
}

@@

\begin{abstract}
Outflows and rotation are two ubiquitous kinematic features in the gas
kinematics of galaxies.
Here we perform Monte Carlo radiative transfer simulations of outflowing
gas with additional solid body rotation to understand how these kinematic
features impact the morphology of the Lyman-$\alpha$ emission line.
We explore a range of neutral Hydrogen optical depth of
$10^5\leq\tauh\leq10^7$, rotational velocity $0\leq \vrot/\kms \leq 100$ and
outflow velocity $0\leq \vout/\kms\leq 50$.  
We find three important consequences of rotation.
First, it introduces a dependency with viewing angle; second it
produces a line broadening and third it increases the flux at the
line's center.
We also test a semi-analytic model that modifies the spectra of
pure outflow simulations and manages to reproduce the radiative
transfer results for the line widths and flux change at the line's
center within a $7\%$ and $50\%$ precision, respectively, for an
optical depth of $\tauh=10^5$ and within $2\%$ and $1\%$ for an
optical depth of $\tauh=10^7$.   
We argue that rotation is a required kinematic feature to take into
account for a correct interpretation of \lya spectra in systems where
rotation is already evident through kinematic maps but also in other
systems, such as Green Pea galaxies, where a strong \lya emission at
the line's center could be explained by rotational kinematics. 
\end{abstract}

@@

Recent advances in instrumentation have revealed the presence of gas
rotation on vastly different physical scales.
For instance, spatially resolved spectra on compact dwarf galaxies
have measured clear signs of gas showing pure rotation kinematics
\citep{2015A&A...577A..21C,2017A&A...600A.125C} and the recent mapping of
high redshift circumgalactic regions have also revealed kinematic
evidence for large scale rotation \citep{2018MNRAS.473.3907A}.
These systems with star formation, neutral gas and low dust
contents naturally show a \lya emission line \citep{PartridgePeebles}
motivating the observational work to phenomenologically link
H$\alpha$ kinematics to \lya spectra \citep[e.g.][]{Herenz2016}.

@@

However, the gas dynamics in Lyman Alpha Emitter (LAE) galaxies are
more complex than pure rotation.
The case for outflows as traced in the \lya line is much better
established.
In many observations the \lya line profile has a single peak
redwards from the line's center, in other cases there is a double peak
but the peak on the red side is stronger
\citep[e.g.][]{2010ApJ...717..289S,Erb14,Trainor16}.   
These features have been readily explained as the consequence of multiple
\lya photon scatterings through a homogeneous outflowing shell of
neutral Hydrogen
\citep{2006A&A...460..397V,Orsi12,2012ApJ...751...29Y,2015ApJ...812..123G}.

An study of the combined effects of outflows and rotation has not been
presented in the literature.
Here we present such a study. 
We investigate a simplified geometrical configuration corresponding to a
spherical gas cloud with symmetrical radial outflows and solid body
rotation. 
We base our modeling on a Monte-Carlo radiative transfer code.
Besides modeling the impact of joint rotation and outflows, we also
check to what extent the analytical model presented by
\cite{Garavito14} to explain the effects of rotation can also be
applied in our case.

@@

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{muse_vs_simulations.pdf}

\caption{\textbf{Comparison between simulations and observations.}  
The observational results correspond to the MUSE-Wide Survey
\citep{2017A&A...606A..12H}. The skewness and standard deviation from
the simulations summarize the results presented in Figures
\ref{fig:standard_deviation} and \ref{fig:skewness}.
  \label{fig:muse}}
\end{figure}

@@

The Monte Carlo code we use (CLARA) \citep{CLARA} follows the propagation of
individual photons through a neutral Hydrogen medium characterized by
its temperature, velocity field and global optical depth.
The code assumes a homogeneous density throughout the simulated
volume.
In the current implementation we neglect the influence of dust.
Our basic model is an spherical distribution of neutral hydrogen,
an approximation commonly used in the literature, as it explains a
wide variety of observational features
\citep{Ahn03,Verhamme06,Dijkstra06}.

The velocity field we use captures both outflows and rotation.
Outflows are described by a Hubble-like radial velocity profile with
the velocity magnitude increasing linearly with the radial
coordinate; the outflows model is fully characterized by $v_{\rm out}$, the
velocity at the sphere's surface.
Rotation follows a solid body rotation profile, which is fully
characterized by $v_{\rm rot}$, the linear velocity at the sphere's surface.

The total velocity field corresponds to the superposition of rotation and
outflows.
The cartesian components take the following form:

@@

\begin{equation}
	v_{z}=\frac{z}{R}\vout,
	\label{eq:vz}
\end{equation}
%
where $x$, $y$ and $z$ are the cartesian position coordinates with the
origin at the sphere's center, $R$ is the radius of the sphere and the
direction of the angular velocity vector corresponds to the \hatk unit vector.

For each model setup we follow $10^5$ individual photons generated at
the center of the sphere at the \lya line's center as they propagate
through the volume and finally scape.
We store the final frequency and propagation direction for each photon
at its last scattering.

As input parameters we use $\tauh=\{10^5, 10^6, 10^7\}$,
$\vout=\{5,25,50\}\;\kms$ and $\vrot=\{0,50,100\}\;\kms$, for a total of
$27$ models with all the possible parameter combinations.
The range of values have some overlap with the expectations from a
galaxy with a total neutral hydrogen mass of $10^8$-$10^9$
$M_{\odot}$.

\cite{Garavito14} presented an analytical model that
accounts for the effects of pure rotation on the
\lya line morphology. 
The basic assumption of their analytical model is that each
differential surface element on the sphere Doppler shifts the photons
that it emits.
In this paper we introduce this ansatz by post-processing the results
of the outflows simulations without rotation.
The frequency of each photon is Doppler shifted as follows

\begin{equation}
x' = x + \frac{\vec{v}_{\rm rot}\cdot \hatk}{\vth}
\label{eq:shift_x}
\end{equation}
%
where $x'$ is the photon's new adimensional frequency, $x$ is the photon's
frequency after being processed by the outflow, \vrot is the 
rotational velocity at the point of escape of the photon, \hatk is
the photon's direction of propagation and \vth is the thermal
velocity of the sphere.

This allows us to produce new \lya spectra and compare them with the
full radiative transfer solution including both outflows and
rotation.

@@

Figure \ref{fig:doppler_shift} summarizes the most important trends.
In the left side, the six panels correspond to $\tau=10^6$ and a viewing angle of
$\theta =90^{\circ}$, that is, perpendicular to the rotation axis of the
galaxy. 
In every panel the thin black line corresponds to the pure outflow
solution, i.e. without rotation. 
From top to bottom we see the effect of increasing the outflow
velocity, which is the expected increasing asymmetry towards the red
peak.

@@

The thick gray line corresponds to the pure outflow solution
with the Doppler boost added to model rotation's influence.
At $\tauh=10^6$ the Doppler boost does a good job at capturing the broad
morphological features introduced by rotation: the angle dependence,
the broadening and the intensity increase at the line's center.

In the right side of Figure \ref{fig:doppler_shift} we show the same results as in the left one, but for a viewing angle of $\theta =
90^{\circ}$, that is parallel to the rotation axis. 
In this case we confirm the result presented by \cite{Garavito14},
namely that pure rotation introduces a strong dependence with 
viewing angle, a trend that we find also holds for rotation mixed with
outflows.

The quality of the results from the Doppler boost improves for higher
\tauh values. 
In the Appendix we show the same plots as Figure \ref{fig:doppler_shift}, there it is
evident that for $\tauh=10^5$ the results are not as good as they are
for $\tauh=10^6$, and that for $\tauh=10^7$ the Doppler boost
provides a remarkable good approximation.

\subsection{Quantitative trends}
\label{sec:quantitative}

After finding the qualitative influence of the different parameters we
move onto a quantitative study.
To do this we summarize the line morphology by four different
characteristics: standard deviation (STD), skewness (SKW), bimodality
(BI) and valley/peak ratio.
These quantities are defined by the following equations \citep{kokoska1999}:

@@

\subsection{\lya continuum leakage or rotation?}

One application of this new model to the current interpretation of \lya
spectra, is that the central ($V=0$) emission of the spectra seen in the
\lya line, is a consequence of the viewing angle of the galaxy and can be
controlled by it. 
Several authors have suggested that this central emission is caused by
radiation that escapes the galaxy without scattering, mostly due to a
clumpy interstellar medium \citep{Hansen06, 2016ApJ...833L..26G}.
Rotation is an alternative that solves this issue.

We note that the relative intensity of the two peaks in the outflowing
spectra is not modified by rotation, that asymmetry is solely
controlled by the outflow velocity.
This means that results that have already derive a typical outflowing
velocity as to match the observational constraints do not have to be
completely revised. 
Only an additional exploration of rotational velocities and viewing
angles needs to be explored.

A recent observational example is provided by
\cite{2017A&A...608L...4R}.
They report optical spectroscopy of a bright lensed galaxy at a
redshift of $z=2.4$. 
The object presents a triple peaked \lya profile that can be thought
as the superposition of two components, a double asymmetric peak and a
narrower symmetric peak around the line's center.
One possible option for the narrow component is that it escaped the
galaxy without substantial interaction with neutral Hydrogen.
We speculate that the central feature could also be due to a purely 
rotating component that can produce such a feature as shown by
\cite{tololo}. The triple peak could be then produced by two main
decopled components: one rotating the other outflowing.

Another interesting example are the \lya lines in Green Pea galaxies \citep{2016ApJ...820..130Y}.
\cite{2017ApJ...844..171Y} showed in a study of 43 Green Pea galaxies
that 2/3 were strong \lya emitters. 
There the \lya escape fraction is effectively estimated from the ratio
of H$\alpha$ to \lya.
Galaxies with large escape fractions also substantial \lya emission at
the line's center. 
We speculate that this feature could help to constrain rotation in
Green Pea galaxies.

\subsection{\lya Kinematic Maps}

Current observational facilities have the capability of spatially
resolving the extent of a LAE.
For instance \cite{Prescott14} presented observational results of a
Doppler shift when taking spectra at two opposite sides of a large
($\approx 80$kpc) LAE.
In more recent work \cite{2018MNRAS.473.3907A} mapped \lya emission
around a quasar.

@@

Although it is a good approximation to think the rotating spectra by a
sum of Doppler shifts, the peak of the spectra is also weighted by the
amount of mass with a given velocity. 
In this toy model the distance between the peaks of the
receding/approaching spectra is close to $\sim 20\ \kms$, which is
less than half of the naively expected value of $2\vrot=100\ \kms$,
due to the fact that only a small fraction of the photons are emitted
at the extreme of the galaxy having the maximum rotational velocity of
$50\ \kms$.

New spectrographs like MUSE should help obtain kinematic information
from LAEs to build velocity maps in Ly$\alpha$. 
This could be a natural extension of the work reported by
\cite{Herenz2016} on the velocity maps of several LARS (Lyman Alpha
Reference Sample) galaxies.  
The interpretation of such data should take into account the insights
and trends we have presented in this paper.

\subsection{Comparison against the MUSE-Wide Survey}

In order to give the reader a feeling of the ranges of observational
data that could be explained by the models presented here, 
we perform a comparison against observational data from the MUSE-Wide
Survey (MWS) \citep{2017A&A...606A..12H}.

MWS provides public access to 831 spectra out of which 237 contain a
detected \lya line as dominant. 
The public dataset reports as the redshift of the galaxy the position
of the dominant peak.
We use this redshift to convert the spectra to a rest-frame and compute
the standard deviation and skewness statistics. 
Given the uncertainty on the galaxy redshift we do not compute the
valley/peak ratio. 
We do not try to perform a full fit to all the spectra
\citep[e.g.][]{2017A&A...608A.139G} and wish only to
know whether the region of parameter space explored in our RT
simulations shows any crossing with the results from the MWS or
shows any sign of atypical features.

Figure \ref{fig:muse} shows the results of this experiment. 
The small white circles correspond to the 217 \lya emitters in the MWS
observations; the large black circles to the results of the
simulations presented in this paper. 
This shows that the systems modeled in this paper actually present an
overlap with statistics derived from observations. 
The region with standard deviation $>200$\kms could also be explained
by systems with larger optical depths than we have used here. 
However, the systems with a positive skewness, that is a line
unbalanced towards the blue could be the sign of inflowing
kinematics.

@@

In this paper we explore, for the first time in the literature,
the results of a model for the emergent \lya line from rotating outflows.
The main results for the model are computed from a Monte-Carlo
radiative transfer simulation and confronted with a simple
semi-analytic ansatz that adds the effects of rotation onto results of
pure outflow kinematics.

To address the first question we posed in the introduction (\emph{what is then
the expected imprint of rotation on a resonant emission line such as
the \lya line?}) we can say that the main effects of rotation on the
\lya line morphology are:

\begin{itemize}
  \item Inducing a dependency on the viewing angle.
  \item Broadening the line.
  \item Increasing the intensity at the line's center.
\end{itemize}

We find that these effects can be quantitatively explained by a
Doppler shift computed as product of quantities at the surface of last
scattering, namely  $\vec{v}_{\rm rot}\cdot \hatk$, where
$\vec{v}_{\rm rot}$ is the velocity due to rotation and \hatk is
the direction of the photon's propagation.  
These conclusions, specially the confirmation of the Doppler Boost as
a good approximation to model solid body rotation, strengthen the
semi-analytic model reported by \cite{Garavito14} where they modeled
the effect of pure rotation on \lya spectra.

Addressing the second question we posed in the introduction (\emph{to what
extent is it possible to constrain rotational kinematics from the \lya
emission line?})  we can say that as an application to observational data our results suggest that
rotation should also be considered as an explanation of the
non-vanishing \lya intensity at the line's center, which could change
the interpretation of recent observational results
\citep[e.g.][]{2017ApJ...844..171Y,2017A&A...608L...4R} that claim
instead that this flux constrains the escape fraction of \lya
radiation.

A more straightforward approach takes into account how the
\lya spectra from two different sides of a galaxy can detect
approaching/receding gas motions.
In that case, one has to be careful and take into account that the
peaks of these two spectra is smaller than the expected value of
$2\times\vrot$ due to the different weights at the surface of the
emitting regions.

We also use data from MUSE to demonstrate to what extent the
outputs of the RT results presented here can be expected to match
hundreds of observed \lya spectra, but a full constraint of rotation
properties certainly require a full Markov Chain Monte Carlo
exploration of parameter space.

Our work confirms that rotation produces features in the \lya line
strong enough that can be expected to be detected in observations and
that should be taken into account into \lya modeling; the Doppler
Boost offers an easy-to-implement approximation to explore such
influence into already existing RT simulations and provide stronger
footing to the interpretation of \lya kinematics maps
\citep[e.g][]{2018MNRAS.473.3907A}.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.49\textwidth]{doppler_shift_logtau5_theta90}
  \end{center}
  \caption{\textbf{Qualitative trends of changing outflow and
      rotational velocity.}
    Same layout as Figure \ref{fig:doppler_shift},
    this time  $\tauh=10^5$ and $\theta=90^\circ$.}
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=0.49\textwidth]{doppler_shift_logtau7_theta90}
  \end{center}
  \caption{\textbf{Qualitative trends of changing outflow and
      rotational velocity.}
    Same layout as Figure \ref{fig:doppler_shift},
    this time  $\tauh=10^7$ and $\theta=90^\circ$.}
\end{figure}
